---
output:
  html_document:
    keep_md: yes
---

[![Build Status](https://travis-ci.org/hadley/devtools.svg?branch=master)](https://travis-ci.org/hadley/devtools)
[![codecov](https://codecov.io/gh/jaredhuling/oem/branch/master/graph/badge.svg)](https://codecov.io/gh/jaredhuling/oem)


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(message = FALSE)
```




## Introduction

The oem package provides estimaton for various penalized linear models using the [Orthogonalizing EM algorithm](http://amstat.tandfonline.com/doi/abs/10.1080/00401706.2015.1054436). Documentation for the package can be found here: [oem site](http://casualinference.org/oem) (still under construction).

Install using the **devtools** package (RcppEigen must be installed first as well):

```r
#devtools::install_github("jaredhuling/oem")
```

or by cloning and building using `R CMD INSTALL`

## Models

### Lasso

```{r}
library(microbenchmark)
library(glmnet)
library(oem)
# compute the full solution path, n > p
set.seed(123)
n <- 1000000
p <- 100
m <- 25
b <- matrix(c(runif(m), rep(0, p - m)))
x <- matrix(rnorm(n * p, sd = 3), n, p)
y <- drop(x %*% b) + rnorm(n)

lambdas = oem(x, y, intercept = TRUE, standardize = FALSE, penalty = "elastic.net")$lambda

microbenchmark(
    "glmnet[lasso]" = {res1 <- glmnet(x, y, thresh = 1e-10, 
                                      standardize = FALSE,
                                      intercept = TRUE,
                                      lambda = lambdas)}, 
    "oem[lasso]"    = {res2 <- oem(x, y,
                                   penalty = "elastic.net",
                                   intercept = TRUE, 
                                   standardize = FALSE,
                                   lambda = lambdas,
                                   tol = 1e-10)},
    times = 5
)

# difference of results
max(abs(coef(res1) - res2$beta[[1]]))

res1 <- glmnet(x, y, thresh = 1e-12, # thresh must be very low for glmnet to be accurate
                                      standardize = FALSE,
                                      intercept = TRUE,
                                      lambda = lambdas)

max(abs(coef(res1) - res2$beta[[1]]))
```

### Nonconvex Penalties

```{r mcp, message = FALSE, warnings = FALSE}
library(sparsenet)
library(ncvreg)
library(plus)
# compute the full solution path, n > p
set.seed(123)
n <- 5000
p <- 200
m <- 25
b <- matrix(c(runif(m, -0.5, 0.5), rep(0, p - m)))
x <- matrix(rnorm(n * p, sd = 3), n, p)
y <- drop(x %*% b) + rnorm(n)

mcp.lam <- oem(x, y, penalty = "mcp",
               gamma = 2, intercept = TRUE, 
               standardize = TRUE,
               nlambda = 200, tol = 1e-10)$lambda

scad.lam <- oem(x, y, penalty = "scad",
               gamma = 4, intercept = TRUE, 
               standardize = TRUE,
               nlambda = 200, tol = 1e-10)$lambda

microbenchmark(
    "sparsenet[mcp]" = {res1 <- sparsenet(x, y, thresh = 1e-10, 
                                          gamma = c(2,3), #sparsenet throws an error 
                                                          #if you only fit 1 value of gamma
                                          nlambda = 200)},
    "oem[mcp]"    = {res2 <- oem(x, y,  
                                 penalty = "mcp",
                                 gamma = 2,
                                 intercept = TRUE, 
                                 standardize = TRUE,
                                 nlambda = 200,
                                 tol = 1e-10)},
    "ncvreg[mcp]"    = {res3 <- ncvreg(x, y,  
                                   penalty = "MCP",
                                   gamma = 2,
                                   lambda = mcp.lam,
                                   eps = 1e-7)}, 
    "plus[mcp]"    = {res4 <- plus(x, y,  
                                   method = "mc+",
                                   gamma = 2,
                                   lam = mcp.lam)},
    "oem[scad]"    = {res5 <- oem(x, y,  
                                 penalty = "scad",
                                 gamma = 4,
                                 intercept = TRUE, 
                                 standardize = TRUE,
                                 nlambda = 200,
                                 tol = 1e-10)},
    "ncvreg[scad]"    = {res6 <- ncvreg(x, y,  
                                   penalty = "SCAD",
                                   gamma = 4,
                                   lambda = scad.lam,
                                   eps = 1e-7)}, 
    "plus[scad]"    = {res7 <- plus(x, y,  
                                   method = "scad",
                                   gamma = 4,
                                   lam = scad.lam)}, 
    times = 5
)


diffs <- array(NA, dim = c(4, 1))
colnames(diffs) <- "abs diff"
rownames(diffs) <- c("MCP:  oem and ncvreg", "SCAD: oem and ncvreg",
                     "MCP:  oem and plus", "SCAD: oem and plus")
diffs[,1] <- c(max(abs(res2$beta[[1]] - res3$beta)), max(abs(res5$beta[[1]] - res6$beta)),
               max(abs(res2$beta[[1]][-1,1:nrow(res4$beta)] - t(res4$beta))),
               max(abs(res5$beta[[1]][-1,1:nrow(res7$beta)] - t(res7$beta))))
diffs

```



### Group Lasso

```{r group, message = FALSE, warnings = FALSE}
library(gglasso)
library(grpreg)
library(grplasso)
# compute the full solution path, n > p
set.seed(123)
n <- 5000
p <- 200
m <- 25
b <- matrix(c(runif(m, -0.5, 0.5), rep(0, p - m)))
x <- matrix(rnorm(n * p, sd = 3), n, p)
y <- drop(x %*% b) + rnorm(n)
groups <- rep(1:floor(p/10), each = 10)

grp.lam <- oem(x, y, penalty = "grp.lasso",
               groups = groups,
               nlambda = 100, tol = 1e-10)$lambda


microbenchmark(
    "gglasso[grp.lasso]" = {res1 <- gglasso(x, y, group = groups, 
                                            lambda = grp.lam, 
                                            intercept = FALSE,
                                            eps = 1e-8)},
    "oem[grp.lasso]"    = {res2 <- oem(x, y,  
                                       groups = groups,
                                       penalty = "grp.lasso",
                                       lambda = grp.lam,
                                       tol = 1e-10)},
    "grplasso[grp.lasso]"    = {res3 <- grplasso(x=x, y=y, 
                                                 index = groups,
                                                 standardize = FALSE, 
                                                 center = FALSE, model = LinReg(), 
                                                 lambda = grp.lam * n * 2, 
                                                 control = grpl.control(trace = 0, tol = 1e-10))}, 
    "grpreg[grp.lasso]"    = {res4 <- grpreg(x, y, group = groups, 
                                             eps = 1e-10, lambda = grp.lam)},
    times = 5
)


diffs <- array(NA, dim = c(2, 1))
colnames(diffs) <- "abs diff"
rownames(diffs) <- c("oem and gglasso", "oem and grplasso")
diffs[,1] <- c(  max(abs(res2$beta[[1]][-1,] - res1$beta)), max(abs(res2$beta[[1]][-1,] - res3$coefficients))  )
diffs

```

#### Bigger Group Lasso Example

```{r grouptwo}
set.seed(123)
n <- 500000
p <- 200
m <- 25
b <- matrix(c(runif(m, -0.5, 0.5), rep(0, p - m)))
x <- matrix(rnorm(n * p, sd = 3), n, p)
y <- drop(x %*% b) + rnorm(n)
groups <- rep(1:floor(p/10), each = 10)

system.time(res <- oem(x, y, penalty = "grp.lasso",
                       groups = groups,
                       standardize = TRUE,
                       intercept = TRUE,
                       nlambda = 100, tol = 1e-10))

# memory usage is out of control here.
# oem uses approximately 1/3 of the memory
system.time(res2 <- grpreg(x, y, group = groups, 
                           eps = 1e-10, lambda = res$lambda))

# I think the standardization is done
# differently for grpreg
max(abs(res$beta[[1]] - res2$beta))
mean(abs(res$beta[[1]] - res2$beta))

```

### Fitting Multiple Penalties

The oem algorithm is quite efficient at fitting multiple penalties simultaneously when p is not too big.

```{r mult, fig.height=6, fig.width = 8, fig.align="center"}
set.seed(123)
n <- 100000
p <- 100
m <- 15
b <- matrix(c(runif(m, -0.25, 0.25), rep(0, p - m)))
x <- matrix(rnorm(n * p, sd = 3), n, p)
y <- drop(x %*% b) + rnorm(n)

microbenchmark(
    "oem[lasso]"    = {res1 <- oem(x, y,
                                   penalty = "elastic.net",
                                   intercept = TRUE, 
                                   standardize = TRUE,
                                   tol = 1e-10)},
    "oem[lasso/mcp/scad/ols]"    = {res2 <- oem(x, y,
                                   penalty = c("elastic.net", "mcp", "scad", "grp.lasso"),
                                   gamma = 4,
                                   groups = rep(1:10, each = 10),
                                   intercept = TRUE, 
                                   standardize = TRUE,
                                   tol = 1e-10)},
    times = 5
)
#png("../mcp_path.png", width = 3000, height = 3000, res = 400);par(mar=c(5.1,5.1,4.1,2.1));plot(res2, which.model = 2, main = "mcp",lwd = 3,cex.axis=2.0, cex.lab=2.0, cex.main=2.0, cex.sub=2.0);dev.off()
#

layout(matrix(1:4, ncol=2, byrow = TRUE))
plot(res2, which.model = 1, main = "lasso", lwd = 2)
plot(res2, which.model = 2, main = "mcp", lwd = 2)
plot(res2, which.model = 3, main = "scad", lwd = 2)
plot(res2, which.model = 4, main = "group lasso", lwd = 2)
```
