---
title: "Usage of the oem Package"
author: "Jared Huling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage of the oem Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`oem` is a package for the estimation of various penalized regression models. The focus of `oem` is to provide high performance computation for big **tall** data. Many applications not only have a large number of variables, but a vast number of observations; `oem` is designed to perform well in these settings.

- Fast computation for big **tall** data
- Efficient computation for computation for multiple penalties simultaneously
- Efficient cross-validation 

## Installation

The simplest way to install `oem` is via the CRAN repositories as the following:

```{r, echo=FALSE, message = FALSE, cache=FALSE}
# install.packages("oem", repos = "http://cran.us.r-project.org")
```

To install the development version, first install the `devtools` package

```{r, message = FALSE, cache=FALSE, eval = FALSE}
install.packages("devtools", repos = "http://cran.us.r-project.org")
```

and then installl `oem` via the `install_github` function
```{r, message = FALSE, cache=FALSE, eval=FALSE}
library(devtools)
install_github("jaredhuling/oem")
```


## Quick Start

First load `oem` 

```{r, echo=FALSE, message = FALSE, cache=FALSE}
library(oem)
```

Simulate some data

```{r, echo=FALSE, message = FALSE, cache=FALSE}
nobs  <- 1e4
nvars <- 100
x <- matrix(rnorm(nobs * nvars), ncol = nvars)
y <- drop(x %*% c(0.5, 0.5, -0.5, -0.5, 1, rep(0, nvars - 5))) + rnorm(nobs, sd = 2)
```


Fit a penalized regression model using the `oem` function

```{r, echo=FALSE, message = FALSE, cache=FALSE}
fit1 <- oem(x = x, y = y, penalty = "lasso")
```

Plot the solution path

```{r, fig.show='hold', fig.width = 7.15, fig.height = 5}
plot(fit1)
```



## Fitting multiple penalties at once

The oem algorithm is well-suited to quickly estimate a solution path for multiple penalties simultaneously if the number of variables is not too large. 

For the group lasso penalty, the `groups` argument must be used. `groups` should be a vector which indicates the group number for each variable.

```{r, echo=FALSE, message = FALSE, cache=FALSE}
fit2 <- oem(x = x, y = y, penalty = c("lasso", "mcp", "grp.lasso"),
            groups = rep(1:20, each = 5))
```

Plot the solution paths for all models

```{r, fig.show='hold', fig.width = 7.15, fig.height = 5}
layout(matrix(1:3, ncol = 3))
plot(fit2, which.model = 1)
plot(fit2, which.model = 2)
plot(fit2, which.model = 3)
```

## Cross Validation

Here we use the `nfolds` argument to specify the number of folds for $k$-fold cross validation

```{r, echo=FALSE, message = FALSE, cache=FALSE}
cvfit1 <- cv.oem(x = x, y = y, penalty = c("lasso", "mcp", "grp.lasso"), 
                 groups = rep(1:20, each = 5), 
                 nfolds = 10)
```

Plot the cross validation results for each model

```{r, fig.show='hold', fig.width = 7.15, fig.height = 4}
layout(matrix(1:3, ncol = 3))
plot(cvfit1, which.model = 1)
plot(cvfit1, which.model = 2)
plot(cvfit1, which.model = 3)
```


