---
title: "Usage of the oem Package"
author: "Jared Huling"
date: "`r Sys.Date()`"
bibliography: oem.bib
output: 
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
        css: oem.css
vignette: >
  %\VignetteIndexEntry{Usage of the oem Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`oem` is a package for the estimation of various penalized regression models using the oem algorithm of [@xiong13]. The focus of `oem` is to provide high performance computation for big **tall** data. Many applications not only have a large number of variables, but a vast number of observations; `oem` is designed to perform well in these settings.

- Fast computation for big **tall** data
- Efficient computation for computation for multiple penalties simultaneously
- Efficient cross-validation 

# Installation

The simplest way to install `oem` is via the CRAN repositories as the following:

```{r, echo=FALSE, message = FALSE, cache=FALSE}
# install.packages("oem", repos = "http://cran.us.r-project.org")
```

To install the development version, first install the `devtools` package

```{r, message = FALSE, cache=FALSE, eval = FALSE}
install.packages("devtools", repos = "http://cran.us.r-project.org")
```

and then installl `oem` via the `install_github` function
```{r, message = FALSE, cache=FALSE, eval=FALSE}
library(devtools)
install_github("jaredhuling/oem")
```


# Quick Start

First load `oem` 

```{r, message = FALSE, cache=FALSE}
library(oem)
```

Simulate some data

```{r, message = FALSE, cache=FALSE}
nobs  <- 1e4
nvars <- 100
x <- matrix(rnorm(nobs * nvars), ncol = nvars)
y <- drop(x %*% c(0.5, 0.5, -0.5, -0.5, 1, rep(0, nvars - 5))) + rnorm(nobs, sd = 4)
```


Fit a penalized regression model using the `oem` function

```{r, message = FALSE, cache=FALSE}
fit1 <- oem(x = x, y = y, penalty = "lasso")
```

Plot the solution path

```{r, fig.show='hold', fig.width = 7.15, fig.height = 5}
plot(fit1)
```


## Available functions

+--------------------+-----------------------------------+
| Function Name      | Functionality                     |
+====================+===================================+
| `oem()`            | Main fitting function             |
+--------------------+-----------------------------------+
| `predict.oemfit()` | Prediction for oem objects        |
+--------------------+-----------------------------------+
| `plot.oemfit()`    | Plotting for oem objects          |
+--------------------+-----------------------------------+
| `logLik.oemfit()`  | log Likelihood for oem objects    |
+--------------------+-----------------------------------+
| `cv.oem()`         | Cross-validation function         |
+--------------------+-----------------------------------+
| `predict.cv.oem()` | Prediction for cv.oem objects     |
+--------------------+-----------------------------------+
| `plot.cv.oem()`    | Plotting for cv.oem objects       |
+--------------------+-----------------------------------+
| `logLik.cv.oem()`  | log Likelihood for cv.oem objects |
+--------------------+-----------------------------------+

# Fitting multiple penalties at once

The oem algorithm is well-suited to quickly estimate a solution path for multiple penalties simultaneously if the number of variables is not too large. 

For the group lasso penalty, the `groups` argument must be used. `groups` should be a vector which indicates the group number for each variable.

```{r, message = FALSE, cache=FALSE}
fit2 <- oem(x = x, y = y, penalty = c("lasso", "mcp", "grp.lasso"),
            groups = rep(1:20, each = 5))
```

Plot the solution paths for all models

```{r, fig.show='hold', fig.width = 7.15, fig.height = 5}
layout(matrix(1:3, ncol = 3))
plot(fit2, which.model = 1)
plot(fit2, which.model = 2)
plot(fit2, which.model = 3)
```

## Timing Comparison

The following is a demonstration of oem's efficiency for computing solutions for tuning parameter paths for multiple
penalties at once

```{r, message = FALSE, cache=FALSE}
nobs  <- 1e5
nvars <- 100
x2 <- matrix(rnorm(nobs * nvars), ncol = nvars)
y2 <- drop(x2 %*% c(0.5, 0.5, -0.5, -0.5, 1, rep(0, nvars - 5))) + rnorm(nobs, sd = 4)

system.time(fit2a <- oem(x = x2, y = y2, penalty = c("grp.lasso"),
                         groups = rep(1:20, each = 5), nlambda = 100L))
system.time(fit2b <- oem(x = x2, y = y2, penalty = c("grp.lasso", "lasso", "mcp", "scad", "elastic.net"),
                         groups = rep(1:20, each = 5), nlambda = 100L))
system.time(fit2c <- oem(x = x2, y = y2, penalty = c("grp.lasso", "lasso", "mcp", "scad", "elastic.net"),
                         groups = rep(1:20, each = 5), nlambda = 500L))
```

# Cross Validation

Here we use the `nfolds` argument to specify the number of folds for $k$-fold cross validation

```{r, message = FALSE, cache=FALSE}
cvfit1 <- cv.oem(x = x, y = y, penalty = c("lasso", "mcp", "grp.lasso"), 
                 groups = rep(1:20, each = 5), 
                 nfolds = 10)
```

Plot the cross validation results for each model

```{r, fig.show='hold', fig.width = 7.15, fig.height = 3.75}
layout(matrix(1:3, ncol = 3))
plot(cvfit1, which.model = 1)
plot(cvfit1, which.model = 2)
plot(cvfit1, which.model = 3)
```

# More Information

For further information about `oem`,
please visit:

  * The oem site:        [casualinference.org/oem](http://casualinference.org/oem)
  * The oem source code: [github.com/jaredhuling/oem](https://github.com/jaredhuling/oem)
  * The oem paper:       [oem paper](http://www.tandfonline.com/doi/abs/10.1080/00401706.2015.1054436)

# References
