---
title: "Usage of the oem Package"
author: "Jared Huling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage of the oem Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`oem` is a package for the estimation of various penalized regression models. The focus of `oem` is to provide high performance computation for big **tall** data. Many applications not only have a large number of variables, but a vast number of observations; `oem` is designed to perform well in these settings.

- Fast computation for big **tall** data
- Efficient computation for computation for multiple penalties simultaneously
- Efficient cross-validation 

## Installation

The simplest way to install `oem` is via the CRAN repositories as the following:

```{r, echo=FALSE}
install.packages("oem")
```

To install the development version, first install the `devtools` package

```{r, echo=FALSE}
install.packages("devtools")
```

and then installl `oem` via the `install_github` function
```{r, echo=FALSE}
devtools::install_github("jaredhuling/oem")
```


## Quick Start

Load `oem` 

```{r, echo=FALSE}
library(oem)
```


The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
